# Forecasting Models and Comparison

## Processing

#### Load Packages

```{r}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(readxl)
library(forecast)
```

#### Load Data

```{r}
train <- read_csv("data/clean_train_monthly.csv")
head(train)
```

#### Convert to Time Series with Only Delay Rate Field

```{r}
train_ts <- train %>%
  select(year, month, delay_rate) %>%
  arrange(year, month)

train_ts <- ts(train_ts$delay_rate,
              start = c(train_ts$year[1], train_ts$month[1]),
              frequency = 12)
```

#### Inspect Delay Rate Time Series in Training Data

```{r}
autoplot(train_ts)
```

#### Decompose Using STL

```{r}
decomp <- stl(train_ts, s.window = "periodic")
autoplot(decomp)
```

#### Test Data Processing

```{r}
test <- read_csv("data/clean_test_monthly.csv")

test_ts <- test %>%
  select(year, month, delay_rate) %>%
  arrange(year, month)

test_ts <- ts(test_ts$delay_rate,
              start = c(test_ts$year[1], test_ts$month[1]),
              frequency = 12)
```

## Model 1: Exponential Smoothing (ETS)

```{r}
ets <- ets(train_ts)
ets
```

```{r}
forecast_ets <- forecast(ets, h = length(test_ts))

autoplot(forecast_ets) +
  autolayer(test_ts, series = "Actual", PI = FALSE) +
  ylab("delay rate") +
  xlab("Time")
```

```{r}
acc_ets <- accuracy(forecast_ets, test_ts)
acc_ets
```

**The ETS model performs well with low bias and \~11% average percentage error, beats the naïve forecast clearly (MASE \< 1, Theil’s U \< 1), and shows slightly better test accuracy than training, suggesting stable forecasting performance.**

## Model 2: SARIMA

```{r}
sarima <- auto.arima(train_ts, seasonal = TRUE)
sarima
```

#### Seasonal ARIMA(1,0,0)(2,1,0)[12]

```{r}
forecast_sarima <- forecast(sarima, h = length(test_ts))

autoplot(forecast_sarima) +
  autolayer(test_ts, series = "Actual", PI = FALSE) +
  ylab("delay rate") +
  xlab("Time")
```

```{r}
acc_sarima <- accuracy(forecast_sarima, test_ts)
acc_sarima
```

-   ME (Mean Error) – forecast bias

    -   Training: -0.0023 -\> model slightly over‑forecasting on training.

    -   Test: 0.0060 -\> slightly under‑forecasting on test.

    -   Values are very close to 0 → little systematic bias, which is good.

-   RMSE (Root Mean Squared Error) – overall error size

    -   Training: 0.0299

    -   Test: 0.0260 -\> smaller than training -\> generalizes reasonably well and is not overfitting.

-   MAE (Mean Absolute Error) – average absolute error

    -   Training: 0.0211

    -   Test: 0.0210 -\> Very similar between training and test → stable model performance

    -   On average, forecasts miss the actual delay rate by about 2.1%

-   MPE (Mean Percentage Error) – percentage bias

    -   Training: -3.09% -\> forecasts about 3% too high on average

    -   Test: 1.61% -\> forecasts about 1.6% too low

-   MAPE (Mean Absolute Percentage Error) – average percent error

    -   Training: 11.86%

    -   Test: 11.07% -\> Forecasts are off by about 11% on average

-   MASE (Mean Absolute Scaled Error) – compared to naïve forecast

    -   Training: 0.637 -\> \< 1 means the model is better than a naïve benchmark

    -   Test: 0.634 -\> about 37% improvement over naïve forecast

-   ACF1 – autocorrelation of residuals at lag 1

    -   Training: -0.061

    -   Test: -0.055

    -   Close to 0, which means residuals are roughly uncorrelated.

    -   Indicates the model captured most temporal structure.

-   Theil’s U Test: 0.503

    -   \< 1 means better than naïve forecasting.

    -   0.50 indicates substantial improvement over a naïve model.

**The SARIMA model appears well calibrated with low bias, stable training vs test performance, and clearly better than a naïve benchmark, with average forecasting error around 11% of the delay rate.**
